---
title: "DATA 607 Project 1"
author: "Jered Ataky"
date: "`r Sys.Date()`"
output: 
  openintro::lab_report: default
  html_document:
    number_sections: yes
---


***Collaborators:*** Dan Rosenfeld, Magnus Skonberg, and Rick Sughrue.



## Overview



<style>
div.aquamarine { background-color:#7fffd4; border-radius: 10px; padding: 5px;}
</style>
<div class = "aquamarine">


In this project, we’re given a text file with chess tournament results where the information has some structure. Our job is to create an R Markdown file that generates a .CSV file (that could for example be imported into a SQL database)
with the following information for all of the players:

Player's Name, Player's State, Total Number of Points, Player's Pre-Rating, and Average Pre Chess Rating of Opponents

For the first player, the information would be:
Gary Hua, ON, 6.0, 1794, 1605

1605 was calculated by using the pre-tournament opponents’ ratings of 1436, 1563, 1600, 1610, 1649, 1663, 1716, and dividing by the total number of games played.

</div> \hfill\break




## Libraries

```{r}
library(stringr)
library(tidyverse)
library(downloader)
library(readr)
library(knitr)
library(kableExtra)
```


## Download text file from GitHub to local machine


```{r import-data}

# Download text file to the local machine

url <- "https://raw.githubusercontent.com/jnataky/DATA606project1/master/tournamentinfo.txt" 
chess_txt <- "tournamentinfo.txt"

downloader::download(url, chess_txt)

getwd()

```


## Read the text file


```{r}

chess <- readLines(chess_txt)

#Get the insight
chess[1:10]

```



## Variables extraction & patterns

<style>
div.aquamarine { background-color:#7fffd4; border-radius: 10px; padding: 5px;}
</style>
<div class = "aquamarine">


*This section is about extraction of all the string variable from chess tournament file provided
following different patterns, then make some transformation and calculation as needed.
Str and unlist will be widely used in here.*

</div> \hfill\break



```{r}

# Patterns

pn <- "\\w+[^USCF|a-z] ?\\w+ \\w+"  # Player's name pattern
ps <- "(?:^|\\W)ON | MI | OH(?:$|\\W)" # Player's state pattern
pp <- "\\d\\.\\d"   # Player's total points pattern


# Extract player's names, player' state, and total point following the above patterns

PlayerName <- unlist(str_extract_all(chess, pn))

PlayerState <- unlist (str_extract_all(chess, ps))

PlayerPoints <- unlist(str_extract_all(chess, pp ))


```



```{r Pre-Rating}

# Pre-rating extraction 
# Patterns

p1 <- "(R:\\s*)(\\d+)"  # Player's pre-rating pattern 1
p2 <- "(\\d+)" # Player's pre-rating pattern 2

# Extraction: this will be done following the two patterns p1 and p2 simultaneously

PreRating <- unlist(str_extract_all(chess, p1))

PreRating <- unlist(str_extract_all(PreRating, p2))

PreRating <- as.numeric(PreRating) # String to numeric

PreRating
```



```{r Average-Rating}
# Extract the average rating

# Patterns

A1 <- "\\|[0-9].*"  # Average rating pattern 1
A2 <- "\\s\\d{1,2}"  # Average rating pattern 2

# String extraction

AveRating <- unlist(str_extract_all(chess, A1))

AveRating <- str_replace_all(AveRating, "\\s{1,2}\\|","0|") # Replace empty (unplayed game) by 0
AveRating <- (str_extract_all(AveRating, A2))


```


```{r Average-Rating-of-the-opponents}

# Put all AveRating into a matrix 
# This will make our further calculation easier 

AveRating_matrix <- matrix(unlist(AveRating),  byrow= TRUE, nrow = length(AveRating))

```

```{r}

# Convert string values of the above matrix to numeric
# Note that apply converts it but chnages the dimension
#Use transform t to bring back the initial dimension

AveRating_matrix_t <- t(apply(AveRating_matrix, 1, as.numeric))

```


```{r}

# Replace 0 values by the NA to calculate further the mean
#by considering only played games
#Note that 0 represented unplayed game


for (r in 1:nrow(AveRating_matrix_t)) {
  
  for (c in 1:ncol(AveRating_matrix_t)) {
    
    if (AveRating_matrix_t[r,c] == 0){
      
      AveRating_matrix_t[r,c] = NA
    } else {
      AveRating_matrix_t[r,c] <- PreRating[AveRating_matrix_t[r,c]]
    }
    
  }
  
  
  
}
  

```



```{r Opponents-Pre-rating-Average}

# Calculate the opponents average using rowMeans

OpponentsAvg <- c(rowMeans(AveRating_matrix_t, na.rm = TRUE))

OpponentsAvg

```

```{r Opponents-Pre-rating-Average-rounded}

# Round the opponents pre-rating average

OpponentsAvg <- round(OpponentsAvg, 0)

OpponentsAvg

```


```{r results-data-frame}

# Construct a data frame for to output the result

results <- data.frame(PlayerName, PlayerState, PlayerPoints, PreRating, OpponentsAvg)

# Rename columns names

colnames(results) <- c("Player's Name", "Player's State", "Total Number of Point", "Player's Pre-Rating", "Opponents Pre-Rating Avg")

```



## Chess Tournament Info

<style>
div.aquamarine { background-color:#7fffd4; border-radius: 10px; padding: 5px;}
</style>
<div class = "aquamarine">


*This section will show the result of the tournament on a table.*

</div> \hfill\break




```{r chess-tournement-table}

results %>%
  kbl(caption = "Chess Tournament Info", align = 'c') %>%
  kable_material(c("striped", "hover")) %>%
  row_spec(0, color = "indigo")

```




## Export to csv


```{r Export-csv}

# Export to csv

write.csv(results, "chessgame.csv", row.names = FALSE )

```



## Extra work: ELO calculation


<style>
div.aquamarine { background-color:#7fffd4; border-radius: 10px; padding: 5px;}
</style>
<div class = "aquamarine">

Calculating the expected score for each player compare to the opponent's average.

Use the above formula to calculate the probability of a player to win the game:

$elo = \frac{1}{1+10^{\frac{B-A}{400}}}$

New rating = rating + 32(score -expected score) (1)

Rewrite (1): N = R + 32(S-elo)

$S = \frac{N-R}{32} + elo$ (2)

Use (2) to calculate the score, S.


With A: Player's Pre-Rating and B: Opponents Pre-Rating Avg


</div> \hfill\break



### ELO calculation

```{r}

# ELO Calculation

results <- results %>%
  mutate(elo_rating = 1/(1 + 10^(((results$`Opponents Pre-Rating Avg`) - (results$`Player's Pre-Rating`))/400 ))) 

# Convert elo to percentage and round the answer by 2 decimals

# results$elo_rating <- 100*results$elo_rating
results$elo_rating <- round(results$elo_rating, 3)

```


### Player Post-Rating extraction

```{r}

# Post-Rating extraction
# Extract all everything first (pre & post rating)

PostRating <- str_extract_all(chess, "(( \\:)|(\\>))?.?\\d{1,}P*\\.?")

# Detect the pattern for post rating

PostRating_detect <- str_detect(unlist(PostRating), "\\>.?\\b\\d{3,4}P?\\b")

# Construct the string of post rating

PostRating_new <-unlist(PostRating)[PostRating_detect]


# Clean the post rating s bit 

PostRating_new <-str_replace_all(PostRating_new, "([>P])","")

# String to numeric
PostRating_new <-as.numeric(PostRating_new)


```

```{r}

# Add post rating to the data frame

results <- results %>%
  mutate(PostRating_new)

```


### Player's actual score

```{r}

actual_score = (((results$`PostRating_new`) - (results$`Player's Pre-Rating`))/32) + results$elo_rating

```


```{r}

# Add columns of actual score

results <- results %>%
  mutate(actual_score) 

# Convert actual score and round the answer by 1 decimal place

results$actual_score <- round(results$actual_score, 1)


```


```{r}

# Arrange players per actual score

new_results <- results %>%
 arrange(desc(actual_score))

# Rename the columns names

colnames(new_results) <- c("Player's Name", "Player's State", "Total Number of Point", "Player's Pre-Rating", "Opponents Pre-Rating Avg", "ELO rating (%)", "Player's Post-Rating", "Actual Score Points")

```


### Tournament table per actual score

```{r chess-tournement-elo-table}

# Result in table.

new_results %>%
  kbl(caption = "Chess Tournament Info per actual score", align = 'c') %>%
  kable_material(c("striped", "hover")) %>%
  footnote(general = "Jacob Alexander Lavalley scored the most points relative to his expected result.") %>%
  row_spec(0, color = "indigo")



```
