---
title: "DATA606 - Data Project"
author: Jered Ataky
date: December 3, 2020
output:
  ioslides_presentation:
    self_contained: true
    widescreen: true
    smaller: true
editor_options: 
  chunk_output_type: console
---


```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
set.seed(2112)
library(statsr)
library(infer)
library(tidyverse)
library(ggplot2)
library(openintro)
library(DATA606)
library(reshape2)
library(psych)
library(reshape2)
options(width=100)
par(mar=c(2.5,1,2,1))
```

# Students Performance in Exams


## Overview

Data is collected by [kaggle](https://www.kaggle.com/spscientist/students-performance-in-exams) to explore and build in a 
web-based data science environment.

There are 1000 observations with 8 variables in this given data set, 
and each case represents a student in the United States. 

It is an observatory study in which: 

- The response variable is mean tests score and is numerical
- The explanatory variable is test preparation course and is categorical.

**Research question:**

"Is the average tests score different from students who
have test preparation course and those who don't ?"


## The data

```{r setup, echo=TRUE, results='hide', warning=FALSE, message=FALSE}

# Load data from Github repository

data <- read.csv("https://raw.githubusercontent.com/jnataky/DATA-607/master/A2_Various_dataset_transformation/students_performance.csv")


data_sub <- data %>%
  select(test.preparation.course, math.score, writing.score, reading.score)

data_clean <- data_sub %>%
  transmute(test.preparation.course, tests_score = (math.score + writing.score + reading.score) / 3)

names(data_clean) <- c("test_prep", "tests_score")

# Check for missing values

sum(is.na(data_clean))

data_final <- data_clean %>%
  transmute(test_prep_course = ifelse(data_clean$test_prep == "completed", "yes", "no"), tests_score)

```


## Exploratory data analysis 


```{r}

test_prep_yes <- data_final %>% filter(test_prep_course == "yes")

test_prep_no <- data_final %>% filter(test_prep_course == "no")

boxplot(test_prep_yes$tests_score, test_prep_no$tests_score,
        names = c("Test score with preparation", "Test score with no preparation"))
```


## Summary statistics

**Description**
```{r}

describe(test_prep_yes$tests_score)

describe(test_prep_no$tests_score)


```

## Summary statistics

**Plot: Distribution of students who had test preparation course**
```{r, warning=FALSE, message=FALSE}

ggplot(test_prep_yes, aes(x = tests_score)) + geom_histogram() +
  ggtitle("Distribution of students who had test preparation course")

```



## Summary statistics

**Plot: Distribution of students who had no test preparation course**
```{r, warning=FALSE, message=FALSE}

ggplot(test_prep_no, aes(x = tests_score)) + geom_histogram() +
  ggtitle("Distribution of students who had no test preparation course")

```


## Inference 

**Conditions:**

- Independence: The sample is made of different students and they are independent each others.
- Random samples:it is a random sample of observations.
- Approximately normal: more than 30 samples, CLT.

**Hypotheses:**

- H0: The average tests score from students who
have test preparation course is the same to those who don't.

- H1: The average tests score from students who
have test preparation course is different to those who don't.

**Significance level:**

$\alpha = 0.05$ 

## Inference 

**Point estimate**

```{r}

point_estimate <- data_final %>%
  specify(tests_score ~ test_prep_course) %>%
  calculate(stat = "diff in means" , order = c("yes", "no")) 

```

**Null distribution**

```{r}

set.seed(1412)

ci_null_dist <- data_final %>%
  specify(tests_score ~ test_prep_course) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 200, type = "permute") %>%
  calculate(stat = "diff in means", order = c("yes", "no"))
```


## Inference 

**Confidence interval**

```{r}
ci_null_dist %>%
  
  get_confidence_interval(point_estimate = point_estimate,
                          
                          level = 0.95,
                          
                          type = "se")
```


We are 95% confident that the difference between the tests score of those who
have test preparation test and those who don't falls in (5.77, 9.50)


## Inference 

**P value**

```{r}

ci_null_dist %>%
  
  get_p_value(obs_stat = point_estimate,
              direction = "two-sided")
```



p_value < 0.05

We reject the null hypothesis. The data provides evidence for the alternative.

Thus, the average tests score from students who
have test preparation course are different to those who don't.

## Inference 

**Visualization**

```{r}
ci_null_dist %>%
  visualise() 

```


## Conclusion

This analysis helps us to understand that test preparation is very important to
the success of students. Students need to consider test preparation course
in order to be successful in exams.

Limitation is in the fact that the test preparation in the analysis can't be 
quantified. This means that the data should at least provide some information on 
the amount of hours of preparation for each student who took the test. 
This will give more general idea on the average time of preparation is needed 
to be successful in exams although we know that the learning curve is 
different from one student to another one.

## References

https://www.kaggle.com/spscientist/students-performance-in-exams