---
title: "Chapter 4 - Distributions of Random Variables"
author: "Jered Ataky"
output:
    pdf_document:
        extra_dependencies: ["geometry", "multicol", "multirow"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
library(DATA606)
```

**Area under the curve, Part I**. (4.1, p. 142) What percent of a standard normal distribution $N(\mu=0, \sigma=1)$ is found in each region? Be sure to draw a graph.

(a) $Z < -1.35$
(b) $Z > 1.48$
(c) $-0.4 < Z < 1.5$
(d) $|Z| > 2$


## Solution 1

```{r}

# (a)

pnorm(-1.35)

```

```{r}
normalPlot(bounds = c(-Inf,-1.35))

```


```{r}

# (b)

1 - pnorm(1.48)

```

```{r}

normalPlot(bounds = c(1.48,Inf))

```




```{r}

# (c)

pnorm(1.5) - pnorm(-0.4) 

```
```{r}

normalPlot(bounds =c(-0.4, 1.5))

```

```{r}

# (d)

1 - pnorm(2) + pnorm(-2)

```

```{r}

# (d)

normalPlot(bounds = c(-2, 2), tails = TRUE)

```

--------------------------------------------------------------------------------

\clearpage

**Triathlon times, Part I** (4.4, p. 142) In triathlons, it is common for racers to be placed into age and gender groups. Friends Leo and Mary both completed the Hermosa Beach Triathlon, where Leo competed in the *Men, Ages 30 - 34* group while Mary competed in the *Women, Ages 25 - 29* group. Leo completed the race in 1:22:28 (4948 seconds), while Mary completed the race in 1:31:53 (5513 seconds). Obviously Leo finished faster, but they are curious about how they did within their respective groups. Can you help them? Here is some information on the performance of their groups:

* The finishing times of the \textit{Men, Ages 30 - 34} group has a mean of 4313 seconds with a standard deviation of 583 seconds.
* The finishing times of the \textit{Women, Ages 25 - 29} group has a mean of 5261 seconds with a standard deviation of 807 seconds.
* The distributions of finishing times for both groups are approximately Normal.

Remember: a better performance corresponds to a faster finish.

(a) Write down the short-hand for these two normal distributions.
(b) What are the Z-scores for Leo's and Mary's finishing times? What do these Z-scores tell you?
(c) Did Leo or Mary rank better in their respective groups? Explain your reasoning.
(d) What percent of the triathletes did Leo finish faster than in his group?
(e) What percent of the triathletes did Mary finish faster than in her group?
(f) If the distributions of finishing times are not nearly normal, would your answers to parts (b) - (e) change? Explain your reasoning.


## Solution 2

(a)

Denote M men's group, and W women's group.

M ~ $N(\mu = 4313 , \sigma = 583)$

W ~ $N(\mu = 5261 , \sigma =  807)$

(b)

```{r}


# Leo' Z score:

z = (4948 - 4313) / 583
z

```
This tells me that Leo is 1.09 standard deviation above the mean.
Thus, Leo's running time is 1 standard deviation above the group mean 
running time.

```{r}

# Mary's z score:

z = (5513 - 5261) / 807
z

```
This tells me that Mary is 0.31 standard deviation above the mean.
Thus, Mary's running time is 0.31 standard deviation above the group mean 
running time.


(c) Leo's z score is higher than Mary's. Meaning Leo is more away from the mean,
his running time is more than the average of his group, thus he didn't perform 
well. That said, Mary, who has the lowest z score, ranked better in her group compared to Leo.


(d)

```{r}

# P(X>4948)

1-pnorm(4948, mean = 4313, sd = 583)
normalPlot(mean = 4313, sd = 583, bounds = c(4948, Inf))

```

Leo finished faster then 13.8 % of the triathletes in his group.

(e)

```{r}

# P(X>5513)

1 - pnorm(5513, mean = 5261, sd = 807)

normalPlot(mean = 5261, sd = 807, bounds = c(5513, Inf))

```
Mary finished faster than 37.7% of the triathletes in her group.



(f) 

For z scores, non normal distributions will be transformed into
sets of z scores, that's it, z scores can be calculated for distributions
that are not normal. Meaning I can still solve (b-c). When it comes to (c-d),
as the distributions are not normal, we can't use the standard normal 
to calculate probability. However, I could calculate the probability  
empirically but will need more information (such as number of runners, 
each individual running time,...) that could help me for calculation. 

--------------------------------------------------------------------------------

\clearpage

**Heights of female college students** Below are heights of 25 female college students.

\[ \stackrel{1}{54}, \stackrel{2}{55}, \stackrel{3}{56}, \stackrel{4}{56}, 
\stackrel{5}{57}, \stackrel{6}{58}, \stackrel{7}{58}, \stackrel{8}{59}, 
\stackrel{9}{60}, \stackrel{10}{60}, \stackrel{11}{60}, \stackrel{12}{61}, 
\stackrel{13}{61}, \stackrel{14}{62}, \stackrel{15}{62}, \stackrel{16}{63}, 
\stackrel{17}{63}, \stackrel{18}{63}, \stackrel{19}{64}, \stackrel{20}{65}, 
\stackrel{21}{65}, \stackrel{22}{67}, \stackrel{23}{67}, \stackrel{24}{69}, 
\stackrel{25}{73} \]

(a) The mean height is 61.52 inches with a standard deviation of 4.58 inches. Use this information to determine if the heights approximately follow the 68-95-99.7% Rule.
(b) Do these data appear to follow a normal distribution? Explain your reasoning using the graphs provided below.


```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold", out.width="50%", fig.height=3}
library(openintro)
heights <- c(54, 55, 56, 56, 57, 58, 58, 59, 60, 60, 60, 61, 
            61, 62, 62, 63, 63, 63, 64, 65, 65, 67, 67, 69, 73)
par(mar=c(3.7,2.2,1,1), las=1, mgp=c(2.5,0.7,0), mfrow = c(1,1), cex.lab = 1.5, cex.axis = 1.5)
histPlot(heights, col = COL[1], xlab = "Heights", ylab = "", probability = TRUE, axes = FALSE, ylim = c(0,0.085))
axis(1)
x <- heights
xfit <- seq(min(x)-5, max(x)+5, length = 400)
yfit <- dnorm(xfit, mean = mean(x), sd = sd(x))
lines(xfit, yfit, col = COL[4], lwd = 2)
par(mar=c(3.7,3.7,1,1), las=1, mgp=c(2.5,0.7,0), mfrow = c(1,1), cex.lab = 1.5, cex.axis = 1.5)
qqnorm(heights, col = COL[1], pch = 19, main = "", axes = FALSE)
axis(1)
axis(2)
qqline(heights, col = COL[1])
```


## Solution 3

(a) I will solve this using normalPlot to show the curve and calculate
the probability with respective conditions of the standard deviation away the 
mean. These values should correspond to 68-95-99.7% to determine if
the heights approximately follow the 68-95-99.7% Rule.
 
 
```{r}
# For 68%: bounds are -/+ 1 sd + mean

normalPlot(mean = 61.52, sd = 4.58, bounds = c(-4.58+61.52, 4.58+61.52))

```

With -/+ 1 sd : 68.3%
```{r}
# For 95%: bounds are -/+ 2 sd + mean

normalPlot(mean = 61.52, sd = 4.58, bounds = c(-4.58*2 + 61.52, 4.58*2 + 61.52))

```

With -/+ 2 sd: 95.4%


```{r}
# For 99.7%: bounds are -/+ 3 sd + mean

normalPlot(mean = 61.52, sd = 4.58, bounds = c(-4.58*3 + 61.52, 4.58*3 + 61.52))

```
With -/+ 3 sd: 99.7%


As we can see from the graphs and values calculated, the heights approximately
follow the 68-95-99.7% Rule.

(b)

```{r}
# Use the DATA606::qqnormsim function

qqnormsim(x)
```

The points in these plots closely follow a diagonal line, that is, it an evident 
to say that these data appear to follow a normal distribution. This is also
confirmed by the histogram and the density of the normal curved given.


--------------------------------------------------------------------------------

\clearpage

**Defective rate.** (4.14, p. 148) A machine that produces a special type of transistor (a component of computers) has a 2% defective rate. The production is considered a random process where each transistor is independent of the others.

(a) What is the probability that the 10th transistor produced is the first with a defect?
(b) What is the probability that the machine produces no defective transistors in a batch of 100?
(c) On average, how many transistors would you expect to be produced before the first with a defect? What is the standard deviation?
(d) Another machine that also produces transistors has a 5% defective rate where each transistor
is produced independent of the others. On average how many transistors would you expect to be produced with this machine before the first with a defect? What is the standard deviation?
(e) Based on your answers to parts (c) and (d), how does increasing the probability of an event affect the mean and standard deviation of the wait time until success?


## Solution 4

(a) 

```{r}

p <- (0.02) * (1 - 0.02)^(10-1)
p

dgeom(10,0.02)

```

the probability that the 10th transistor produced is the first with a defect
is 0.01667496

(b)
```{r}

dbinom(0,100,0.02)

```
The probability that the machine produces no defective transistors 
in a batch of 100 is 0.1326196

(c)

```{r}

geom_mean <- 1 / 0.02
geom_mean

```

```{r}

geom_std <- sqrt((1 - 0.02) / (0.02)^2)
geom_std

```
On average, I would expect 50 transistors to be produced before the first 
with a defect with a the standard deviation of 49.49747

(d)

```{r}

geom_mean2 <- 1 / 0.05
geom_mean2

```

```{r}

geom_std2 <- sqrt((1 - 0.05) / (0.05)^2)
geom_std2

```

On average, I would expect 20 transistors to be produced before the first 
with a defect with a the standard deviation of 19.49359

(e) Increasing the probability of an event decreases the mean and standard
deviation of the wait time until success.


--------------------------------------------------------------------------------

\clearpage

**Male children.** While it is often assumed that the probabilities of having a boy or a girl are the same, the actual probability of having a boy is slightly higher at 0.51. Suppose a couple plans to have 3 kids.

(a) Use the binomial model to calculate the probability that two of them will be boys.
(b) Write out all possible orderings of 3 children, 2 of whom are boys. Use these scenarios to calculate the same probability from part (a) but using the addition rule for disjoint outcomes. Confirm that your answers from parts (a) and (b) match.
(c) If we wanted to calculate the probability that a couple who plans to have 8 kids will have 3 boys, briefly describe why the approach from part (b) would be more tedious than the approach from part (a).


## Solution 5

(a)

```{r}
# We can use the formula choose(3,2) * (0.51)^2 * (1-0.51)^(3-2) or simply:

dbinom(2,3,0.51)

```
The probability that two of them will be boys is 0.382347


(b)

Denote B boy, and G girl.

- B B G (0.51) * (0.51) * (0.49) = 0.127449
- B G B (0.51) * (0.49) * (0.51) = 0.127449
- G B B (0.49) * (0.51) * (0.51) = 0.127499

Adding them: 3 * (0.127449) = 0.382347

Both answers from parts (a) and (b) match.

(c) 

The approach from part (b) would be more tedious than the approach from
part (b) because we will have a lot more options (56) to write with a couple who
plans to have 8 kids will have 3 boys from approach in part (b) 
while it is just one line of code from approach in part (a).



--------------------------------------------------------------------------------

\clearpage

**Serving in volleyball.** (4.30, p. 162) A not-so-skilled volleyball player has a 15% chance of making the serve, which involves hitting the ball so it passes over the net on a trajectory such that it will land in the opposing team's court. Suppose that her serves are independent of each other.

(a) What is the probability that on the 10th try she will make her 3rd successful serve?
(b) Suppose she has made two successful serves in nine attempts. What is the probability that her 10th serve will be successful?
(c) Even though parts (a) and (b) discuss the same scenario, the probabilities you calculated should be different. Can you explain the reason for this discrepancy?


## Solution 6

(a)

```{r}

# It is a negative binomial distribution: choose(9,2) * (0.15)^3 * (1 - 0.15)^7
# or simply:

dnbinom(7, 3, 0.15)

```

The probability that on the 10th try she will make her 3rd successful serve
is 0.03895012 (~3.895%)


(b)

The probability that her 10th serve will be successful is 0.15 (15%)

(c) 

Her serves are independent of each other, and in part (b) 
it is the probability of a single success serve which is given 
(A not-so-skilled volleyball player has a 15% chance of making the serve) while
in part (a) it is a combination of 10 serves to get to 3 successes so the 
probability depends on that combination. 