---
title: "Chapter 5 - Foundations for Inference"
author: "Jered Ataky"
output:
  pdf_document:
    extra_dependencies:
    - geometry
    - multicol
    - multirow
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DATA606)
```

**Heights of adults.** (7.7, p. 260) Researchers studying anthropometry collected body girth measurements and skeletal diameter measurements, as well as age, weight, height and gender, for 507 physically active individuals. The histogram below shows the sample distribution of heights in centimeters.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3}
library(openintro)
data(bdims)
par(mar=c(3.7,2.5,0.5,0.5), las=1, mgp=c(2.5,0.7,0), cex.lab = 1.5)
histPlot(bdims$hgt, col = COL[1], xlab = "Height", ylab = "")
```

(a) What is the point estimate for the average height of active individuals? What about the median?
(b) What is the point estimate for the standard deviation of the heights of active individuals? What about the IQR?
(c) Is a person who is 1m 80cm (180 cm) tall considered unusually tall? And is a person who is 1m 55cm (155cm) considered unusually short? Explain your reasoning.
(d) The researchers take another random sample of physically active individuals. Would you expect the mean and the standard deviation of this new sample to be the ones given above? Explain your reasoning.
(e) The sample means obtained are point estimates for the mean height of all active individuals, if the sample of individuals is equivalent to a simple random sample. What measure do we use to quantify the variability of such an estimate (Hint: recall that $SD_x = \frac{\sigma}{\sqrt{n}}$)? Compute this quantity using the data from the original sample under the condition that the data are a simple random sample.


## Solution 1

a.
```{r}

mean(bdims$hgt)
```
The mean is 171.1


b.
```{r}
# Std

sd(bdims$hgt)
```
The SD is 9.4

```{r}
#IQR

IQR(bdims$hgt)

```
IQR = 14

c.

Note that be considered unusual, one should be more than 2 SD away from the mean
in either direction (|z| < 2)
```{r}
# Create an interval for usual height (where 95% of observation will fit in)

mean1 <- 171.1
sd1 <- 9.4

# Lower boundary lb, and upper boundary hb
lb <- mean1 - (2 * sd1)
lb
hb <- mean1 + (2 * sd1)
hb

```

The usual boundary is I = (152.3, 189.9)

* 180 is lower than hb and is within I. 
Thus, the person who is 180cm is not considered unusually tall.

* 155 is greater than lb and is within I.
Therefore, the person who is 155cm is not considered unusually short.


d. 

From central limit theorem (CLT), I will say that the sampling distribution of 
the sample mean is approximately normal, meaning if the researchers take 
another sample that would follow CLT, therefore I would expect the researchers
to obtain similar values but not necessary the same.

e.

That is the standard error.

```{r}
n <- 507
SE <- (sd1) / sqrt(n)
SE

```

The standard error, SE = 0.417


--------------------------------------------------------------------------------

\clearpage

**Thanksgiving spending, Part I.** The 2009 holiday retail season, which kicked off on November 27, 2009 (the day after Thanksgiving), had been marked by somewhat lower self-reported consumer spending than was seen during the comparable period in 2008. To get an estimate of consumer spending, 436 randomly sampled American adults were surveyed. Daily consumer spending for the six-day period after Thanksgiving, spanning the Black Friday weekend and Cyber Monday, averaged $84.71. A 95% confidence interval based on this sample is ($80.31, $89.11). Determine whether the following statements are true or false, and explain your reasoning.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3}
library(openintro)
data("thanksgiving_spend")
par(mar=c(3.7,2.2,0.5,0.5), las=1, mgp=c(2.5,0.7,0), cex.lab = 1.5)
histPlot(thanksgiving_spend$spending, col = COL[1], xlab = "Spending", ylab = "")
```

(a) We are 95% confident that the average spending of these 436 American adults is between $80.31 and $89.11.
(b) This confidence interval is not valid since the distribution of spending in the sample is right skewed.
(c) 95% of random samples have a sample mean between $80.31 and $89.11.
(d) We are 95% confident that the average spending of all American adults is between $80.31 and $89.11.
(e) A 90% confidence interval would be narrower than the 95% confidence interval since we don't need to be as sure about our estimate.
(f) In order to decrease the margin of error of a 95% confidence interval to a third of what it is now, we would need to use a sample 3 times larger.
(g) The margin of error is 4.4.


## Solution 2

n = 436
mean = 84.71
95% CI = (80.31, 89.11)

(a) False. Confidence interval is plausible range of values 
for the population parameter not the sample statistic.

(b) False. The confidence interval can be calculated for a right skewed
given that n is greater to 30 as well. 
(Although we can find cases where samples don't contai the population true mean)

(c) False. Again, Confidence interval is plausible range of values 
for the population parameter not the sample statistic.

(d). True. Confidence interval is plausible range of values 
for the population parameter

(e) True. As the confidence level decreases, the plausible range of values 
become narrower and we can miss some values, and don't want to be that sure
with a narrow interval of plausible values.

(f) False. There is a sqrt on the formula, we need to use 
a sample 9 times larger

(g) True. Margin of error, MOE = |80.31-84.71| = |89.11-84.71| = 4.4



--------------------------------------------------------------------------------

\clearpage

**Gifted children, Part I.** Researchers investigating characteristics of gifted children col- lected data from schools in a large city on a random sample of thirty-six children who were identified as gifted children soon after they reached the age of four. The following histogram shows the dis- tribution of the ages (in months) at which these children first counted to 10 successfully. Also provided are some sample statistics.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3}
library(openintro)
data(gifted)
par(mar=c(3.7,2.2,0.5,0.5), las=1, mgp=c(2.5,0.7,0), cex.lab = 1.5)
histPlot(gifted$count, col = COL[1], 
         xlab = "Age child first counted to 10 (in months)", ylab = "", 
         axes = FALSE)
axis(1)
axis(2, at = c(0,3,6))
```

\begin{tabular}{r | l}
n   & 36 \\
min & 21 \\
mean    & 30.69 \\
sd  & 4.31 \\
max & 39 
\end{tabular}

(a) Are conditions for inference satisfied?
(b) Suppose you read online that children first count to 10 successfully when they are 32 months old, on average. Perform a hypothesis test to evaluate if these data provide convincing evidence that the average age at which gifted children fist count to 10 successfully is less than the general average of 32 months. Use a significance level of 0.10.
(c) Interpret the p-value in context of the hypothesis test and the data.
(d) Calculate a 90% confidence interval for the average age at which gifted children first count to 10 successfully.
(e) Do your results from the hypothesis test and the confidence interval agree? Explain.

## Solution 3


(a) Yes they are met.

**Independence:** random sampling

**Sample size:** Success and failure at least 10 %


(b)

```{r}
#: H0: pop_mean = 32, H1: pop_mean < 32

n3 <- 36
min3 <- 21
mean3 <- 30.69
sd3 <- 4.31
max3 <- 39
pop_mean <- 32

SE <- sd3 / (sqrt(n3))

z <- (mean3 - pop_mean) / SE

normalPlot(bounds = c(-Inf, z))
 

```
p = 0.034 < 0.10: Reject the null hypothesis.
The data provide the evidence for the alternative hypothesis.


(c) That means obtaining a sample statistics at least as extreme as the 
observed data is extremely unlikely to happen just by chance.

(d) 

```{r}

z3 <- qnorm(0.05) # Z*
MOE <- abs(z3 * SE)  # Margin of error

li <- mean3 - MOE  # lower boundary
li
ui <- mean3 + MOE  # upper boundary
ui

```
The confidence interval is CI = (29.51, 31.87)

(e) 

Yes they both agree. 
Confidence interval calculated above doesn't not contained 
the null value so we reject it the same result we have with hypothesis test.
  
--------------------------------------------------------------------------------

\clearpage

**Gifted children, Part II.** Exercise above describes a study on gifted children. In this study, along with variables on the children, the researchers also collected data on the mother's and father's IQ of the 36 randomly sampled gifted children. The histogram below shows the distribution of mother's IQ. Also provided are some sample statistics.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3}
library(openintro)
data(gifted)
par(mar=c(3.7,2.2,0.5,0.5), las=1, mgp=c(2.5,0.7,0), cex.lab = 1.5)
histPlot(gifted$motheriq, col = COL[1], 
         xlab = "Mother's IQ", ylab = "", axes = FALSE)
axis(1)
axis(2, at = c(0,4,8,12))
```

\begin{tabular}{r | l}
n   & 36 \\
min & 101 \\
mean    & 118.2 \\
sd  & 6.5 \\
max & 131 
\end{tabular}

(a) Perform a hypothesis test to evaluate if these data provide convincing evidence that the average IQ of mothers of gifted children is different than the average IQ for the population at large, which is 100. Use a significance level of 0.10.
(b) Calculate a 90% confidence interval for the average IQ of mothers of gifted children.
(c) Do your results from the hypothesis test and the confidence interval agree? Explain.


## Solution 4

(a)

$H0: \mu = 100; H1: \mu \ne 100$;


```{r}

n4 <- 36
min4 <- 101
mean4 <- 118.2
sd4 <- 6.5
max4 <- 131

pop_mean4 <- 100

SE4 <- sd4 / (sqrt(n4))

z4 <- (mean4 - pop_mean4) / SE4
p4 <- 2 * pnorm(- abs(z4))


```

p-value is ~ 0 < 0.10. We reject the null hypothesis.
The data provide the evidence for the alternative hypothesis.

(b)

```{r}

z4 <- qnorm(0.05) # Z*
MOE4 <- abs(z4 * SE4)  # Margin of error

li4 <- mean4 - MOE4  # lower boundary
li4
ui4 <- mean4 + MOE4  # upper boundary
ui4

```

The confidence interval is CI = (116.42, 119.98)
which doesn't include the null value.

(c)

They agree. Confidence interval calculated above doesn't not contained 
the null value so we reject it the same result we have with hypothesis test.



--------------------------------------------------------------------------------

\clearpage

**CLT.** Define the term "sampling distribution" of the mean, and describe how the shape, center, and spread of the sampling distribution of the mean change as sample size increases.

## Solution 5

Where items are sampled, the sampling distribution of the mean is the mean
of the population from that.In other words, the sampling distribution shows the 
distribution of n samples from a given population.
As the sample size increases, the accuracy improves then the spread goes down,
and the shape approaches a normal distribution. That's said, 
the mean becomes a center of spread and more samples.

--------------------------------------------------------------------------------

\clearpage

**CFLBs.** A manufacturer of compact fluorescent light bulbs advertises that the distribution of the lifespans of these light bulbs is nearly normal with a mean of 9,000 hours and a standard deviation of 1,000 hours.

(a) What is the probability that a randomly chosen light bulb lasts more than 10,500 hours?
(b) Describe the distribution of the mean lifespan of 15 light bulbs.
(c) What is the probability that the mean lifespan of 15 randomly chosen light bulbs is more than 10,500 hours?
(d) Sketch the two distributions (population and sampling) on the same scale.
(e) Could you estimate the probabilities from parts (a) and (c) if the lifespans of light bulbs had a skewed distribution?


## Solution 6

(a)

```{r}
mean6 <- 9000
sd6 <- 1000

normalPlot(mean = mean6, sd = sd6, bounds = c(10500, Inf))

```

The probability that a randomly chosen light bulb lasts more than 10,500 hours
is 0.0668 or 6.68%


(b)

```{r}
n6 <- 15

d6<- rnorm(n6, mean = mean6, sd = sd6)
hist(d6, probability = TRUE)
x <- 0:15000
y <- dnorm(x =x, mean = mean(d6), sd = sd(d6))
lines(x = x, y = y, col ="red")
abline(v = mean(d6), col ="tomato")


```

The distribution of the mean lifespan of 15 light bulbs is normal.

(c)

```{r}

SE6 <- sd6 / sqrt(n6)

normalPlot(mean = mean6, sd = SE6, bounds = c(10500, Inf))



```

The probability that the mean lifespan of 15 randomly chosen light bulbs 
is more than 10,500 hours is approximativily 0.

(d)

let consider the population np = 12000 light bulbs
```{r}

n6 <- 15
np <- 10000

d6<- rnorm(n6, mean = mean6, sd = sd6)
hist(d6, probability = TRUE, xlim = c(6000, 12000))


```

```{r}

dp<- rnorm(np, mean = mean6, sd = sd6)
hist(dp, probability = TRUE, xlim = c(6000, 12000))

```


(e) 

We wouldn't estimate the probabilities. 
For part (a), the distribution is not normal so can't use the z table. 
For part (c) however, we would have had a possibility to do so if the sample 
size was large. Per CLT, the distribution could be close to normal but here 
the sample is just 15, so we wouldn't.
  

--------------------------------------------------------------------------------

\clearpage

**Same observation, different sample size.** Suppose you conduct a hypothesis test based on a sample where the sample size is n = 50, and arrive at a p-value of 0.08. You then refer back to your notes and discover that you made a careless mistake, the sample size should have been n = 500. Will your p-value increase, decrease, or stay the same? Explain.

## Solution 7

$z = \frac{\mu - x}{SE}$

The sample size affect SE calculation.

When n increases, SE decreases, so z increases.

p-value decreases.



