---
title: "Lab 1: Intro to R"
author: "Jered Ataky"
date: "9/6/2020"
output: openintro::lab_report
---

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
```

### Exercise 1
Extract counts of girls baptized

```{r view-girls-counts}
arbuthnot$girls
```

**Data Visualization**

```{r}

ggplot(data = arbuthnot, aes(x = year, y = girls)) + geom_point()


```



***Plot with line***

```{r}

ggplot(data = arbuthnot, aes(x=year, y=girls)) + geom_line()

```



### Exercise 2

Apparent trend in the number of girls baptized

```{r trend-girls}

# Adding new variable to the data frame

arbuthnot <- arbuthnot %>%
  mutate(total = boys + girls)

# Plot for the trend

ggplot(data = arbuthnot, aes(x=year, y= total)) + geom_line()
```
```{r}
# Create two new variables on data frame

arbuthnot <- arbuthnot %>%
  mutate(boy_to_girl_ratio = boys / girls)

arbuthnot <- arbuthnot %>%
  mutate(boy_ratio = boys /total)

```


### Exercise 3

***Plot proportion of boys over time***

```{r plot-prop-boys-arbuthnot}
ggplot(data = arbuthnot, aes(x = year, y = boy_ratio)) + geom_line()

```

***Add a data frame with logical data***
```{r}
arbuthnot <- arbuthnot %>%
  mutate(more_boys = boys > girls)

```

***Find min and max values of columns***
```{r}

arbuthnot %>%
  summarise(min = min(boys), max = max(boys))
```
### Exercise 4

Explore new data frame

```{r dim-present}
# Show years included in the data set
present$year
```
The present data set includes years from 1940 to 2002.

```{r}
# Dimensions of data frame
dim.data.frame(present)

```

```{r}
# Retrieve columns names
colnames(present)

```
### Exercise 5

Compare Arbuthnot's and present data set counts.

```{r count-compare}
# Arbuthnot's dim

arbuthnot$year
```

```{r}

# Arbuthnot dim

dim.data.frame(arbuthnot)
```

Arbuthnot data set includes year from 1629 to 1710.
Thus, Arbuthnot data set has 81 different years
while The present data set has 62 different years.

They are not of a similar magnitude.
Arbuthnot data set is 19 years longer than present data set.


### Exercise 6

Proportion of boys over time.

```{r plot-prop-boys-present}
# Create 2 new variables in present data set

present <- present %>%
  mutate(total = boys + girls)

present <- present %>%
  mutate(boy_ratio = boys / total)

# Plot proportion of boys born over time for present data set

ggplot(data = present, aes(x = year, y = boy_ratio)) + geom_line()

```


The proportion of born over time declines from 1940 to late 60's then increases
a bit between late 60's to mid 70's, then start decreasing again.
Overall, we can say that the proportion of boys over time decreases.


***Boys to girls ratio in US***

```{r more_boys}

present <- present %>%
  mutate(more_boys = boys > girls)

present$more_boys
```


As we can see, that is true. Arbuthnot's observation about boys being born
in greater proportion than girls holds up in the U.S as well.


```{r prop boys-girls present}

present <- present %>%
  mutate(boy_to_girl_ratio = boys / girls)

ggplot(data = present, aes(x = year, y = boy_to_girl_ratio)) + geom_line()


```



### Exercise 7

Most number of birth in U.S

```{r find-max-total}
present %>%
  arrange(desc(total))

```


As we can see from the above, the most total number of births in U.S was in 1961.